---
title: "NYPD Analysis"
output:
  html_document:
    df_print: paged
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning and Analysis

```{r echo = T, results = 'hide', message=FALSE}
#load required libraries
library(tidyverse)
library(skimr)
library(ggplot2)
library(caret)
```

```{r}
# File name
file_name = "NYPD_Shooting_Incident_Data__Historic_.csv"

data = read_csv(file_name)

# Generate a detailed summary
data_summary_detailed <- skim(data)

# Print the detailed summary
print(data_summary_detailed)
```

```{r}
for (col in c("LOCATION_DESC", "PERP_AGE_GROUP", "PERP_RACE", "PERP_SEX")) {
  unique_vals <- unique(data[[col]])
  print(paste("Unique values in", col, ":", toString(unique_vals)))
}
```
Since LOC_OF_OCCUR_DESC, LOC_CLASSFCTN_DESC columns complete rate is 0.0628, we are going to remove them

Also, for LOCATION_DESC, replace empty values with NONE, 

For PERP_AGE_GROUP, PERP_RACE replace empty values with UNKNOWN,

For PERP_SEX replace empty values with U

We will also want to create a new column based on OCCUR_TIME column and then only choose columns that we want 

```{r}
# Replace empty values in the LOCATION_DESC column with 'NONE'
data$LOCATION_DESC <- ifelse(is.na(data$LOCATION_DESC), 'NONE', data$LOCATION_DESC)

# Replace empty values in the PERP_AGE_GROUP column with 'UNKNOWN'
data$PERP_AGE_GROUP <- ifelse(is.na(data$PERP_AGE_GROUP), 'UNKNOWN', data$PERP_AGE_GROUP)
data$PERP_AGE_GROUP <- ifelse(data$PERP_AGE_GROUP == "(null)", 'UNKNOWN', data$PERP_AGE_GROUP)

# Replace empty values in the PERP_RACE column with 'UNKNOWN'
data$PERP_RACE <- ifelse(is.na(data$PERP_RACE), 'U', data$PERP_RACE)
data$PERP_RACE <- ifelse(data$PERP_RACE == '(null)', 'U', data$PERP_RACE)

# Replace empty values in the PERP_SEX column with 'U'
data$PERP_SEX <- ifelse(is.na(data$PERP_SEX), 'U', data$PERP_SEX)
data$PERP_SEX <- ifelse(data$PERP_SEX == '(null)', 'U', data$PERP_SEX)

# Create new column based on OCCUR_TIME column
data <- data %>%
  mutate(OCCUR_TIME_PERIOD = case_when(
    between(hour(OCCUR_TIME), 0, 6) ~ "Early Morning",
    between(hour(OCCUR_TIME), 6, 12) ~ "Morning",
    between(hour(OCCUR_TIME), 12, 18) ~ "Afternoon",
    TRUE ~ "Night"
  ))

# Choose columns that we want, also "X_COORD_CD", "Y_COORD_CD", "Latitude", "Longitude", "Lon_Lat" seem not to be useful, so I am not going to choose them too
selected_columns <- c("INCIDENT_KEY", "OCCUR_TIME_PERIOD", "BORO", "PRECINCT",
                      "LOCATION_DESC", "STATISTICAL_MURDER_FLAG",
                      "PERP_AGE_GROUP", "PERP_SEX", "PERP_RACE",
                      "VIC_AGE_GROUP", "VIC_SEX", "VIC_RACE")

data <- data %>%
  select(all_of(selected_columns))
```
We are then create a simple model which is logistic regression in this case

```{r}
# Select relevant columns for modeling
model_data <- data %>%
  select(BORO, OCCUR_TIME_PERIOD, STATISTICAL_MURDER_FLAG)

# Convert categorical variables to factors
model_data <- model_data %>%
  mutate_at(vars(BORO, OCCUR_TIME_PERIOD),
            factor)

# Split data into training and testing sets
set.seed(1)  # For reproducibility
train_index <- createDataPartition(model_data$STATISTICAL_MURDER_FLAG, p = 0.7, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

# Build a logistic regression model
model <- glm(STATISTICAL_MURDER_FLAG ~ ., data = train_data, family = "binomial")

# Make predictions on the test set
predictions <- predict(model, newdata = test_data, type = "response")

# Convert predicted probabilities to binary predictions
predicted_labels <- ifelse(predictions > 0.5, 1, 0)

# Evaluate the model
confusion_matrix <- table(predicted_labels, test_data$STATISTICAL_MURDER_FLAG)
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)

# Print model
summary(model)

# Print the confusion matrix and accuracy
print(confusion_matrix)
print(paste("Accuracy:", accuracy))
```
The accuracy of the model is 0.8 which is not bad.

Only two variables, namely BOROMANHATTAN and OCCUR_TIME_PERIODMorning, have demonstrated statistical significance. These two variables exhibit positive coefficients, suggesting an increased likelihood of murder occurrences in the Manhattan borough and during the morning time period.

## Visualization

```{r pressure, echo=FALSE}
# Create a grouped bar plot
boro_time_plot <- ggplot(data, aes(x = BORO, fill = OCCUR_TIME_PERIOD)) +
  geom_bar(position = "dodge") +
  labs(title = "Number of Cases in Each Borough by Occurrence Time Period",
       x = "Borough",
       y = "Number of Cases") +
  theme_minimal() +
  scale_fill_manual(values = c("Early Morning" = "skyblue", "Morning" = "orange", "Afternoon" = "gray","Night" = "black"))

# Print the plot
print(boro_time_plot)

# Filter data for incidents that occurred in Brooklyn
brooklyn_data <- data[data$BORO == "BROOKLYN", ]

# Create a plot for victim race in Brooklyn
race_plot <- ggplot(brooklyn_data, aes(x = VIC_RACE, fill = VIC_RACE)) +
  geom_bar() +
  labs(title = "Victim Race Distribution in Incidents in Brooklyn",
       x = "Victim Race",
       y = "Number of Cases") +
  theme_minimal()+
  theme(axis.text.x = element_blank())

# Print the plot
print(race_plot)
```

From the plot, we can see that most of the cases are in Brooklyn, and the victim race are mostly Black

Also these are a few bias that can affect our analysis:

Geographic Bias: If certain areas are over represented or underrepresented in the data, the analysis might not reflect the overall patterns accurately.

Missing Data Bias: If important variables are missing or incomplete, the analysis might be biased.